The Evolution of Phenotypically Invariant Genetic Networks

### 2018-08-18 review-responses.tex

This may be because I am more familiar with Fisher's geometric model than
systems biology models, but it seems to me that the first part is more novel,
and should therefore deserve your attention first, should you follow my
suggestion of publishing the two parts separately.
\end{quote}

\begin{point}{Format:}
    Please add line numbers to your manuscript; it is straightforward
with the lineno package in LaTeX.  Please also do not increase text width too
much, as this decreases legibility (LaTeX's default settings already optimize
the number of characters per line).  Finally, be careful about not confusing
citep and citet (a lot of missing parentheses around citations in the second part).
\end{point}

\reply{
}

\begin{point}{all figures:}
    Please ensure that all axes legends and labels are big enough
and not distorted (same of arrow labels on the diagrams). Please also make sure
that the lines are thick enough to be visible.
\end{point}

@@

\reply{
}

\begin{point}{4.}
    Your ms does not simply assume that there is a large ``neutral network''
underlying a phenotype, but suggests a mechanistic underpinning to create this
neutral space. Since much of the appeal of the first part of the paper is
connected to this fact, I'd like to see some more in-depth discussion.
\textit{(suggestions to follow)}
\end{point}

\reply{
}

\begin{point}{(mechanistic models)}
    You should probably mention somewhere that mechanistic models of neutral
  spaces exist in the (quite different) context of RNA folding (papers by Hofacker, Fontana, etal)
\end{point}

@@

\begin{point}{(mechanistic models)}
    You gain a lot of ``neutrality'' by the assumption that the dimension of the
  ``krypotype'' is larger than the dimension of the trait. If the trait is the
expression of a gene at the top of some pathway, this is certainly true.
However, if we include further traits that are affected by the same genes
(i.e., pleiotropy) this is no longer clear. The space of ``all traits under
selection'' of an organism is awfully high-dimensional and pleiotropy is
wide-spread (even if we do not believe in ``omnigenic'' models). It is not clear
to me from what kind of data we could learn more about these dimensions and I
do not expect an answer in your ms, but the issue deserves more discussion.
Currently, ``pleiotropy'' is not even mentioned in the ms.   
\end{point}

\reply{
    We have added discussion of this in the Discussion. \revref
}

\begin{point}{Introduction:}
    ``Genotypes encoding identical phenotypes can even persist stably within a species'' - if there is population structure, I suppose.
\end{point}

\reply{
    Yes, that's right. \revref
}

\begin{point}{Introduction:}
    ``It is not a new observation that there is often more than one way to do the same thing, and that this may lead to speciation'' - Isn't this what people would call a neutral Dobzhansky-Muller incompatibility?
\end{point}

@@

\begin{point}{Results:}
    ``Of course, neither of these are necessarily true for real systems'' - delete ``necessarily''?
\end{point}

\reply{
    This sentence has been removed.
}

\begin{point}{Eq. 3:}
    What does a coordinate change mean biologically? In our JC, we discussed for a while whether this is just a different parametrization of the exact same biological thing (like transformation to principle components). I think I now understand that it is more, but this should be explained better.
\end{point}

\reply{
    XXX add discussion of equivalent oscillator examples XXX
}

### 2018-08-18 kalman_walk_paper.tex
minor changes

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

>>>>

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\paragraph{The dimensionality of trait space} \revpoint{1}{8}
We have focused in examples on single traits (where the phenotype is one-dimensional),
but phenotypes under selection are often high-dimensional,
and different traits often share a genetic basis.
If the phenotype and kryptotype are of the same dimension,
then we do not in general have degrees of freedom, i.e., neutral directions.
The true ``dimensionality of trait space'' has been debated at some length in the literature
\citep{trait_space}.
Whatever it is, however, it seems unlikely 
that it would approach the dimensionality of kryptotype space,
which is something related to the number of genes.
Even under universal pleiotropy \citep{universal_pleiotropy,omnigenic},
XXX
